CL-PFU: A suite of R packages for energy conversion chain analysis

### 2023-11-04 paper.Rmd
Add a sentence noting that MW must be converted to RUVY
Update notation around IEA EWEB data.

<!-- Define some macros for later use -->
[link-cran]:             https://cran.r-project.org
[link-ecctools]:         https://github.com/earamendia/ECCTools
[link-eroitools]:        https://github.com/earamendia/EROITools
[link-iea]:              https://www.iea.org
[link-ieatools]:         https://github.com/MatthewHeun/IEATools
[link-ieaweb]:           https://www.iea.org/data-and-statistics/data-product/world-energy-balances
[link-ilo]:              https://www.ilo.org/
[link-fao]:              https://www.fao.org/
[link-matsbyname]:       https://github.com/MatthewHeun/matsbyname
[link-matsindf]:         https://github.com/MatthewHeun/matsindf
[link-mwtools]:          https://github.com/EnergyEconomyDecoupling/MWTools
[link-pfusetup]:         https://github.com/EnergyEconomyDecoupling/PFUSetup
[link-pfudatabase]:      https://github.com/EnergyEconomyDecoupling/PFUDatabase
[link-pfuaggdatabase]:   https://github.com/EnergyEconomyDecoupling/PFUAggDatabase
[link-pfupipelinetools]: https://github.com/EnergyEconomyDecoupling/PFUPipelineTools
[link-pins]:             https://pins.rstudio.com
[link-rclabels]:         https://github.com/MatthewHeun/RCLabels
[link-recca]:            https://github.com/MatthewHeun/Recca
[link-targets]:          https://docs.ropensci.org/targets
[link-tidyverse]:        https://www.tidyverse.org

@@

Historically, SEA practitioners have
analyzed the ECCs of individual countries
using linked spreadsheets,
often starting with primary- and final-stage data 
from the [International Energy Agency][link-iea]'s (IEA's)
[world energy balances][link-ieaweb] (WEBs)
[@Ayres:2003ec; @Serrenho:2014aa; @Brockway:2014aa; @Brockway:2015aa]. 
Data were stored in varying and inconsistent formats.
An early SEA database [@De-Stercke:2014]
estimated energy efficiencies 
of end-use machines using the economic status of countries,
thereby precluding use in energy-economy studies.

@@

The most important design decision for the suite of `R` packages
involved data format.
We authors are among those who
developed the Physical Supply-Use Table (PSUT) framework
[@Rocco:2016; @Guevara:2017; @Heun:2018; @Aramendia:2022tv],
an adaptation of economic Input-Output (IO) analysis [@Miller:2009]
to describe energy flows through society.
The Physical Supply-Use Table (PSUT) framework 
is a matrix-based approach that tracks energy flows 
("products" in PSUT terminology) 
from resource extraction to
processing stages
("industries" in PSUT terminology)
and, ultimately, 
to final demand
in a way that is compatible with the [IEA][link-iea]'s
[WEB][link-ieaweb] data
[@International-Energy-Agency:2023aa].
We chose the PSUT framework as the data format for the new `R` packages, 
because it succinctly describes an entire ECC
for one country and one year
via a set of six matrices
(the "RUVY" matrices).
See the following table.

@@

Broadly speaking, 
four calculation steps are required to create a CL-PFU database. 
Each step is assisted by functions in one or more of the new `R` packages.
First, the IEA's primary- and final-stage 
WEB data must be converted
to RUVY matrices for each country and year, 
a task accomplished by functions in the [IEATools][link-ieatools] package [@Heun-IEATools:2023]. 
Second, human and animal muscle work must be calculated from 
[International Labor Organization][link-ilo] (ILO) and
[Food and Agriculture Organization][link-fao] (FAO) data,
following the methodology of @Steenwyk:2022ww, 
using functions in the [MWTools][link-mwtools] package [@Marshall:2023ab]. 
Third, the IEA's primary- and final-stage ECC data 
must be extended to the useful stage by
(a) allocating final stage energy to end-use machines and
(b) multiplying allocated final energy by the 
final-to-useful efficiency of each machine.
This task is accomplished by functions in the
[Recca][link-recca] [@Heun-Recca:2023]
package.
Fourth, ECCs must be converted from energy terms to exergy terms. 
This step is also performed by functions in the [Recca][link-recca] package.

@@

| Package | Purpose |
|:--------|:--------|
| [IEATools][link-ieatools]             | Converts [IEA WEB][link-ieaweb] data to RUVY matrices in [matsindf][link-matsindf] data frames |
| [MWTools][link-mwtools]               | Converts [ILO][link-ilo] and [FAO][link-fao] data to RUVY matrices of human and animal muscle work in [matsindf][link-matsindf] data frames |
| [Recca][link-recca]                   | Performs `R` energy conversion chain analysis |
| [PFUSetup][link-pfusetup]             | Identifies input and output locations for the [PFUDatabase][link-pfudatabase] and [PFUAggDatabase][link-pfuaggdatabase] pipelines |
| [PFUPipelineTools][link-pfupipelinetools] | Provides basic functionality for all PFU pipelines |
| [PFUDatabase][link-pfudatabase]       | Provides a [targets][link-targets] pipeline to create a data frame of RUVY matrices |
| [PFUAggDatabase][link-pfuaggdatabase] | Provides a [targets][link-targets] pipeline to aggregate RUVY matrices |

Input data for the PFU database can be found in @Marshall:2023aa.
Access to the full CL-PFU database can be obtained via correspondence 
with author [PEB](mailto:P.E.Brockway@leeds.ac.uk)^[Because the PFU Database contains
primary and final energy [IEA WEB][link-ieaweb] data, 
use of the PFU Database
is restricted to those who have access to [IEA][link-iea] [WEB][link-ieaweb] data.].
If desired, researchers with access to the 
[IEA][link-iea]'s [WEB][link-ieaweb] 
data can create their own CL-PFU database 
following the example below.

@@

```{r, eval=FALSE}
# (1) Purchase, download, and store IEA WEB data in the correct location.
#     Sample data for two countries (Ghana and South Africa) 
#     and two years (1971 and 2000) are included (with permission)
#     at the correct location in the repository for this paper.
# (2) Download FAO and ILO data for muscle work calculations
#     via the included script.
#     This download step is needed only once, and
#     downloaded data are stored in the repository 
#     for this paper.
source(file.path("ExampleFolder", "DownloadScripts", "download_mw_data.R"))
```

>>>>

<!-- Define some macros for later use -->
[link-cran]:             https://cran.r-project.org
[link-ecctools]:         https://github.com/earamendia/ECCTools
[link-eroitools]:        https://github.com/earamendia/EROITools
[link-iea]:              https://www.iea.org
[link-ieatools]:         https://github.com/MatthewHeun/IEATools
[link-ieaeweb]:          https://www.iea.org/data-and-statistics/data-product/world-energy-balances
[link-ilo]:              https://www.ilo.org/
[link-fao]:              https://www.fao.org/
[link-matsbyname]:       https://github.com/MatthewHeun/matsbyname
[link-matsindf]:         https://github.com/MatthewHeun/matsindf
[link-mwtools]:          https://github.com/EnergyEconomyDecoupling/MWTools
[link-pfusetup]:         https://github.com/EnergyEconomyDecoupling/PFUSetup
[link-pfudatabase]:      https://github.com/EnergyEconomyDecoupling/PFUDatabase
[link-pfuaggdatabase]:   https://github.com/EnergyEconomyDecoupling/PFUAggDatabase
[link-pfupipelinetools]: https://github.com/EnergyEconomyDecoupling/PFUPipelineTools
[link-pins]:             https://pins.rstudio.com
[link-rclabels]:         https://github.com/MatthewHeun/RCLabels
[link-recca]:            https://github.com/MatthewHeun/Recca
[link-targets]:          https://docs.ropensci.org/targets
[link-tidyverse]:        https://www.tidyverse.org

@@

Historically, SEA practitioners have
analyzed the ECCs of individual countries
using linked spreadsheets,
often starting with primary- and final-stage data 
from the [International Energy Agency][link-iea]'s (IEA's)
[extended world energy balances][link-ieaeweb] (EWEBs)
[@Ayres:2003ec; @Serrenho:2014aa; @Brockway:2014aa; @Brockway:2015aa]. 
Data were stored in varying and inconsistent formats.
An early SEA database [@De-Stercke:2014]
estimated energy efficiencies 
of end-use machines using the economic status of countries,
thereby precluding use in energy-economy studies.

@@

The most important design decision for the suite of `R` packages
involved data format.
We authors are among those who
developed the Physical Supply-Use Table (PSUT) framework
[@Rocco:2016; @Guevara:2017; @Heun:2018; @Aramendia:2022tv],
an adaptation of economic Input-Output (IO) analysis [@Miller:2009]
to describe energy flows through society.
The Physical Supply-Use Table (PSUT) framework 
is a matrix-based approach that tracks energy flows 
("products" in PSUT terminology) 
from resource extraction to
processing stages
("industries" in PSUT terminology)
and, ultimately, 
to final demand
in a way that is compatible with the [IEA][link-iea]'s
[EWEB][link-ieaeweb] data
[@International-Energy-Agency:2023aa].
We chose the PSUT framework as the data format for the new `R` packages, 
because it succinctly describes an entire ECC
for one country and one year
via a set of six matrices
(the "RUVY" matrices).
See the following table.

@@

Broadly speaking, 
four calculation steps are required to create a CL-PFU database. 
Each step is assisted by functions in one or more of the new `R` packages.
First, the IEA's primary- and final-stage 
EWEB data must be converted
to RUVY matrices for each country and year, 
a task accomplished by functions in the [IEATools][link-ieatools] package [@Heun-IEATools:2023]. 
Second, human and animal muscle work must be calculated from 
[International Labor Organization][link-ilo] (ILO) and
[Food and Agriculture Organization][link-fao] (FAO) data,
following the methodology of @Steenwyk:2022ww, 
using functions in the [MWTools][link-mwtools] package [@Marshall:2023ab]. 
The muscle work data must also be converted to RUVY matrices.
Third, the IEA's primary- and final-stage ECC data 
must be extended to the useful stage by
(a) allocating final stage energy to end-use machines and
(b) multiplying allocated final energy by the 
final-to-useful efficiency of each machine.
This task is accomplished by functions in the
[Recca][link-recca] [@Heun-Recca:2023]
package.
Fourth, ECCs must be converted from energy terms to exergy terms. 
This step is also performed by functions in the [Recca][link-recca] package.

@@

| Package | Purpose |
|:--------|:--------|
| [IEATools][link-ieatools]             | Converts [IEA EWEB][link-ieaeweb] data to RUVY matrices in [matsindf][link-matsindf] data frames |
| [MWTools][link-mwtools]               | Converts [ILO][link-ilo] and [FAO][link-fao] data to RUVY matrices of human and animal muscle work in [matsindf][link-matsindf] data frames |
| [Recca][link-recca]                   | Performs `R` energy conversion chain analysis |
| [PFUSetup][link-pfusetup]             | Identifies input and output locations for the [PFUDatabase][link-pfudatabase] and [PFUAggDatabase][link-pfuaggdatabase] pipelines |
| [PFUPipelineTools][link-pfupipelinetools] | Provides basic functionality for all PFU pipelines |
| [PFUDatabase][link-pfudatabase]       | Provides a [targets][link-targets] pipeline to create a data frame of RUVY matrices |
| [PFUAggDatabase][link-pfuaggdatabase] | Provides a [targets][link-targets] pipeline to aggregate RUVY matrices |

Input data for the PFU database can be found in @Marshall:2023aa.
Access to the full CL-PFU database can be obtained via correspondence 
with author [PEB](mailto:P.E.Brockway@leeds.ac.uk)^[Because the PFU Database contains
primary and final energy [IEA EWEB][link-ieaeweb] data, 
use of the PFU Database
is restricted to those who have access to [IEA][link-iea] [EWEB][link-ieaeweb] data.].
If desired, researchers with access to the 
[IEA][link-iea]'s [EWEB][link-ieaeweb] 
data can create their own CL-PFU database 
following the example below.

@@

```{r, eval=FALSE}
# (1) Purchase, download, and store IEA EWEB data in the correct location.
#     Sample data for two countries (Ghana and South Africa) 
#     and two years (1971 and 2000) are included (with permission)
#     at the correct location in the repository for this paper.
# (2) Download FAO and ILO data for muscle work calculations
#     via the included script.
#     This download step is needed only once, and
#     downloaded data are stored in the repository 
#     for this paper.
source(file.path("ExampleFolder", "DownloadScripts", "download_mw_data.R"))
```
